---
title: "Financial econometrics: H1 - Basic Statistical and Time-Series Concepts"
output: H1_notebook
---

Using R, simulate n = 100 observations from the stochastic process in Question 4, which is reproduced below for your convenience: $$X_t = Z_{t−1} +2Z_t +Z_{t+1}$$
where ($$Z_t$$) ∼ i.i.d. N(0, $$\sigma^2$$) by imposing a Normal/Gaussian distribution on $$Z_t$$.

a) Calculate the simulated autocorrelation coefficients, $$\hat{\rho}_X (h)$$, for h = 1, · · · , 20 for a
single run/replication. Briefly comment on the results, and compare and contrast the results obtained in this part with the analytical results obtained in Question 4.

```{r}
library(tseries)
set.seed(0)
lags <- 0:20 # array of lag values
rho_X <- c(c(1, 2/3, 1/6), rep(0,times=(21-3))) # my theoretical calculation of autocorrelation rho_X

get_X_t = function(len) {
  n = len
  Z_t <- rnorm(n+2)
  X_t <- Z_t[1:n]+2*Z_t[2:(n+1)]+Z_t[3:(n+2)]
  return(X_t)
}
```

```{r}
n=100
X_t = get_X_t(n)
acf(X_t,lag.max=20, type="correlation")
points(lags, rho_X)
```
It's clear the autocorrelation drops to close to zero (expected for h>=3) and then somewhat randomly oscillates about zero, with the magnitude of oscillation consistent with strong white noise. 


(b) Repeat the simulation exercise in part (a) for n=1,000 still for a single run/replication. Critically compare and contrast the results obtained in this part with those in part (a).
```{r}
n=1000
X_t = get_X_t(n)
acf(X_t,lag.max=20, type="correlation")
points(lags, rho_X)
```
Similar qualitative behavior, with a better match to the analytically calculated ACF values. Oscillation for lag > 3 is smaller, consistent with strong white noise.

(c) Repeat the simulation exercise in part (a) for n=100 but, this time, for 1000 runs/replications. Critically compare and contrast the results obtained in this part with those in part (a)
and also in part (b).
```{r}
n=100

acf_mat <- matrix(NA, nrow=1000, ncol=21)

for(i in 1:1000) {
  X_t = get_X_t(n)
  acf_mat[i,] = as.vector(acf(X_t,lag.max=20,plot=FALSE,type="correlation")$acf) #[2:21]
}

acf_mean = colMeans(acf_mat)
plot(0:20, acf_mean, type="h", xlim=c(0, 20), ylim=c(min(acf_mean), max(acf_mean)),
     xlab="Lag", ylab="Autocorrelation", main="Mean ACF")
abline(h=0, lty=2)
points(lags, rho_X)
```

Similar pattern as before, now with much smaller oscillation since we're averaging over many trials.
